Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-11-09 ms.tex
Refine discussion

\section{Discussion}
% General problem. An example of solution. Key features

@@

The proposed \textsc{vdd} system breaks from current norms for scientific data publishing in that it builds directly off hosts supporting \texttt{git} repositories, specifically that run by the commercial company \texttt{GitHub}.

The design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).

GitHub - commercial

Integrate with DOIs

Future technology changes

(ideas) Also the OKFN data package idea http://data.okfn.org might be worth merging in here, especially as we can fill most of the required data fields from the DESCRIPTION file.

(Matt says drop) Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost. The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.

A central feature of this concept is that the versioned data is maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for the collection, curation, and distribution of the data. The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably. We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past. Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.

While our first attempt \textsc{vdd} system, \texttt{datastorr} is currently being used, there remain several challenges that we anticipate will be overcome as both technology and scientific norms change. Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data. Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.

(idea) Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

Something about DOI: easy to integrate.

(old) Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years. At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}). Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}. It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project. \texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Fig. \ref{fig:semantic} and ][]{Nielsen-2017}. This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.

% I'm unclear on the precice benefits of dois here (and actually in
% general to a point); some further clarification might be useful. So
% far as I know they are useful because (a) journal urls are terrible
% and inconsistent (b) because everyone has them. I don't really know
% why else you'd need one. Especially in our case we offer no promise
% of persistence (deleting the github repo would break the doi and we
% could update the repo to include different data trivially). So what
% is being added? Our retrival system does not use the doi at all!

% WKC: hmmmm not sure we've nailed this yet, see
% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819

% We have constant drama at work because the different spreadsheet
% writers we use across the platforms quote files inconsistently so
% every save is a complete rewrite of the file. It's totally
% impossible to do any collaborative editing....
% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project \emph{maintainers} to be reviewed and potentially included in the project. This process is called submitting a ``Pull request''. One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.

 % This section conflates git and github more than is useful. It
 % might be worth setting up github as the explicit platform here or
 % dialing back the language a bit.

%% Git-LFS tries to step around this by storing large files separately
% from git's general machinery. Git-annex did the same. But they
% suffer from some of the other issues with git which are that it's a
% bit of a pain for (especially novice) users to do nontrivial things
% with. I _always_ have to look up how to compare two versions across
% branches. And there's no simple way of loading in two versions of
% the data into R without ending up with two working copies or doing
% some nasty "git show" call.

% I really think that something like CKAN (though it doesn't support
% holding data if I remember correctly) could totally be a complete
% solution to the sorts of issues that we have here.

\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.  Indeed, as with every technology the best available approach is certain to change: there are currently a number of promising technologies---CKAN, rdrop2 and Git-LFS---that may facilitate even better \textsc{vdd} systems in the future.  There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.

In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data that are cheap and relatively easy to set up.

@@

%\begin{table}[h!]
%\centering
%\caption{Alternative platforms for minting \textsc{doi}s from \texttt{GitHub} repositories.}
%{\footnotesize
%\vspace{1cm}
%  \begin{tabular}{p{4cm}p{8cm}}
%  \hline
%  \textbf{Provider} & \textbf{Features} \\ \hline
%  \texttt{Zenodo}: \smurl{zenodo.org} & Automated \textsc{doi} for each \texttt{GitHub} release \\
%    & Funded and maintained by CERN and OpenAIRE for long term preservation \\
%    & At present, can only archive files embedded directly in the \texttt{git} repo, not produced by it as outputs\\
%  \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
%    & Allows you to upload additional files, e.g. datasets built by code \\
%    & Integrated with several journals, e.g. Nature family, \textsc{PNAS} \\
%  \hline
%  \end{tabular}
%  }
%\label{tab:doi_minting}
%\end{table}

%\newpage

>>>>

\section{Discussion}

@@

A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

An important concern for any data delivery system is the stability and reliability of the system. In the short term, users want minimal downtime, high speed, and seamless operation. As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves. In the long term, scientists desire long-term preservation of datasets, software, and published papers. To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release. Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data. Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies---such as CKAN, OKFN or Git-LFS---facilitate even better \textsc{vdd} systems in the future.  There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.

In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system. By adopting best practices from software engineering, we believe a relatively easy, cheap, and scalable solution for serving versioned data now exists.

@@
